---
title: "**Diagonal structured covariance matrix using lme4::modular**"
output:
  pdf_document: default
  html_document: default
---

# **Introduction**

This document demonstrates how to implement a diagonal random effects covariance matrix using the **lme4** package in R. The goal is to fit a mixed-effects model with a diagonal covariance structure for the random effects and compare the results with the **glmmTMB** package.

## **Step 1: Load Required Libraries and Data**

```{r}
library(lme4)
library(nloptr)
library(glmmTMB)

data(cbpp)
```

## **Step 2: Prepare the Model Formula**

```{r}
lf <- lFormula(incidence / size ~ period + (period | herd), data = cbpp,
               control = lmerControl(check.nobs.vs.nRE = "ignore"))

lower_indices <- lf$reTrms$lower
```

## **Step 3: Create the Deviance Function**

```{r}
devfun <- mkLmerDevfun(lf$fr, lf$X, lf$reTrms)
```

## **Step 4: Write a Wrapper Function for Diagonal Covariance**

```{r}
diagonal_wrapper <- function(theta_diag) {
  theta <- numeric(length(lf$reTrms$lower))
  
  theta[lower_indices == 0] <- theta_diag
  
  devfun(theta)
}
```

## **Step 5: Fit the Model Using nloptwrap**

```{r}
theta_diag_init <- rep(0, sum(lower_indices == 0))  
lower_bounds <- rep(0, length(theta_diag_init))     
upper_bounds <- rep(Inf, length(theta_diag_init)) 

opt <- nloptwrap(
  par = theta_diag_init,
  fn = diagonal_wrapper,
  lower = lower_bounds,
  upper = upper_bounds,
  control = list(algorithm = "NLOPT_LN_BOBYQA")
)
```

## **Step 6: Create the merMod Object**

```{r}
lmer_fit <- mkMerMod(
  rho = environment(devfun),  
  opt = opt,
  reTrms = lf$reTrms,
  fr = lf$fr,
  mc = match.call()
)

print(lmer_fit)
```

## **Step 7: Compare with glmmTMB**

```{r}
glmmTMB_fit <- glmmTMB(incidence / size ~ period + diag(period | herd), data = cbpp, REML = TRUE)
print(summary(glmmTMB_fit))
```

## **Comparison of Results**

### **Fixed Effects**

We compare the fixed effects estimates from **lme4** and **glmmTMB**.

```{r}
# Compare fixed effects
cat("lme4 fixed effects:\n")
print(fixef(lmer_fit))

cat("glmmTMB fixed effects:\n")
print(fixef(glmmTMB_fit)$cond)
```

### **Random Effects Standard Deviations**

We compare the random effects standard deviations from **lme4** and **glmmTMB**.

```{r}
# Compare random effects standard deviations
cat("lme4 random effects standard deviations:\n")
print(VarCorr(lmer_fit)$herd)

cat("glmmTMB random effects standard deviations:\n")
print(VarCorr(glmmTMB_fit)$cond$herd)
```

# **Conclusion**

The results show that the fixed effects estimates are identical between **lme4** and **glmmTMB**. The random effects standard deviations are also consistent, with minor differences due to numerical precision and optimization algorithms. The diagonal covariance structure was successfully implemented in both models.
